---
title: "Analyzing microbial interactions with microbimixr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing microbial interactions with microbimixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(microbimixr)
library(ggplot2)
library(patchwork)
```

A common experimental design in microbiology is to mix together two different 
microbes (different genotypes of bacteria, for example) then measure how their 
behavior depends on mix frequency. How do they act differently together than when they're on their own? How does mixing affect their survival and reproductive success (their fitness)?

microbimixr is an R package for analyzing microbial mix experiments. It helps researchers get the most out of their data by providing convenience functions to 
calculate and plot fitness measures that are: 

* Robust and quantitatively comparable across different species and types of 
  interaction
* Useful for both individual and group-centered approaches to social interaction
* Well-suited to statistical analysis of effect sizes and confidence intervals

<!-- 
* Meaningful for both kin and multilevel selection theories of social evolution
-->

The basic workflow for microbimixr is: 

1. Start from a data frame of observed microbial abundances
2. Calculate fitness measures with `calculate_mix_fitness()`
3. Inspect a diagnostic plot of fitness measures with `plot_mix_fitness()`
4. Plot measures of interest with `plot_strain_fitness()`, 
  `plot_total_group_fitness()`, and/or `plot_within_group_fitness()`
5. Fit statistical models of your choosing and plot them with the data


## Starting data

The starting point for microbimixr is a data frame that has the observed abundance of each microbe at the beginning of the experiments and at the end. The data could be any combination of strain abundance (like cell counts or cell densities), strain frequency (the fraction of the total), or total group abundance (all strains added together)---anything that's enough to define the absolute abundance of each strain. Each row of the data frame should have all the data for a single observational unit: one experimental replicate with one strain combination, one mix frequency, one combination of experimental treatments, and so on.

For example, we can look at some datasets included in the package. This is from a small experiment mixing a wild-type strain of *Myxococcus* bacteria with an experimentally-evolved cheater strain and letting them make multicellular fruiting bodies together (smith _et al._&nbsp;2010). The abundance quantities here are the number of cells from each strain before and after development. 

```{r data-smith2010}
head(data_smith_2010)
```

Here's another example. This is from an experiment that grew an antibiotic-sensitive strain of *E.&nbsp;coli* together with an antibiotic-resistant strain that detoxifies its local environment (Yurtsev *et al.*&nbsp;2013). The abundance quantities are total cell density (in units of OD~600~) and the fraction of cells that are the resistant strain (from flow cytometry). The data include several experimental treatments with different concentrations of antibiotic (`ampicillin`) and different amounts of culture growth (`dilution`). 

```{r data-Yurtsev2013}
head(data_Yurtsev_2013)
```

microbimixr works with most types of microbial abundance data. 


## Calculate fitness

There are lots of quantities you can calculate to measure microbial survival and reproductive success that work well enough for a particular biological system and research question. But only a few fitness measures are robust enough to be quantitatively meaningful across different species and different types of interaction. microbimixr makes it easy to calculate them. 

The microbimixr function to calculate fitness is `calculate_mix_fitness()`. It accepts an input data frame of microbial abundances and a character vector describing what the data columns are. It returns a data frame with several calculated measures of fitness and two measures of mix frequency. For example: 

```{r fitness-smith2010}
fitness_smith_2010 <- calculate_mix_fitness(
	data = data_smith_2010, 
	var_names = c(
		initial_number_A = "initial_cells_evolved",
		initial_number_B = "initial_cells_ancestral",
		final_number_A = "final_spores_evolved",
		final_number_B = "final_spores_ancestral", 
		name_A = "GVB206.3", 
		name_B = "GJV10"
	)
)
head(fitness_smith_2010)
```

To include data columns in the fitness output that describe experimental treatments, experimental block, and so on, use the `keep` argument. For the Yurtsev data, for example, we want to keep antibiotic concentration and culture dilution: 

```{r fitness-Yurtsev2013}
fitness_Yurtsev_2013 <- calculate_mix_fitness(
	data_Yurtsev_2013, 
	var_names = c(
		initial_number_total = "OD_initial",
		initial_fraction_A = "fraction_resistant_initial",
		final_number_total = "OD_final",
		final_fraction_A = "fraction_resistant_final",
		name_A = "AmpR",
		name_B = "AmpS"
	), 
	keep = c("ampicillin", "dilution")
)
head(fitness_Yurtsev_2013)
```

Note the warning messages here. `calculate_mix_fitness()` will warn you if any of the data are not biologically meaningful. There cannot be a negative number of cells, for example, and fractions are bounded at zero and one. Some of the Yurtsev data include negative fractions---likely a minor artifact of correcting for background fluorescence in flow cytometry. 

<!-- 
* Explain `name_A` and `name_B` arguments?
* They make explicit which is which in the output itself
* Can be column or label
* Column example: Madgwick data
-->


## Fitness math: what and why

`calculate_mix_fitness()` calculates four fitness outcomes from the input data:

* `fitness_A` : Fitness of strain A
* `fitness_B` : Fitness of strain B
* `fitness_total` : Total-group fitness (strains A and B combined)
* `fitness_ratio_A_B` : Within-group relative fitness measured as `fitness_A / fitness_B`

Some interaction effects are easier to visualize, analyze, and interpret with strain-focused fitness outcomes (`fitness_A` and `fitness_B`), while others are better suited to group-focused outcomes (`fitness_total` and `fitness_ratio_A_B`). So microbimixr includes both approaches by default. 

The actual calculations are very simple. If the initial abundance of strain A (measured as the number or density of cells or virions) is $n_A$ and its final abundance is $n'_A$, then its fitness is $w_A = n'_A/n_A$. Similarly, the fitness of strain B is $w_B = n'_B/n_B$. Total group fitness is $W = (n'_A + n'_B)/(n_A + n_B)$ and the within-group fitness ratio is $w_A/w_B$. 

These values are unscaled Wrightian fitness measured over the entire time period between the initial and final abundances. They measure fold change in absolute abundance. For example: if a bacterial strain's cell density increases 100-fold, its fitness will be $w = 100$. If it decreases to 10% of its initial density, its fitness will be $w = 0.1$.

These fitness measures are robust across microbial species and types of interaction, make fitness effects quantitatively comparable across systems, and can be meaningfully incorporated into theoretical models of microbial social evolution (smith & Inglis 2021). They are usually best visualized and analyzed over $\log_{10}$ scales. 

`calculate_mix_fitness()` also calculates two measures of mixing frequency: 

* `initial_fraction_A` : Initial frequency of strain A measured as the fraction of all cells or virions
* `initial_ratio_A_B` : Initial ratio of strain A to strain B

Microbial interactions may be more relevant or more linear over one scale than the other, so microbimixr includes both by default. `initial_fraction_A` is calculated as $n_A / (n_A + n_B)$ and runs from 0 to 1. `initial_ratio_A_B` is calculated as $n_A / n_B$. It runs from negative infinity to positive infinity and is best analyzed over $\log_{10}$ scales. 


## Compare fitness measures

To quickly see which fitness measures are most useful for analyzing your dataset, use `plot_mix_fitness()`. 
<!-- 
`plot_mix_fitness()` produces a diagnostic overview of the various fitness effects in a dataset.  
To quickly see an overview of the various fitness effects in a dataset, use `plot_mix_fitness()`. 
-->
It draws a combined figure with strain fitness, total group fitness, and relative within-group fitness plotted against two measures of mix frequency.

```{r compare-smith, fig.width = 6.25, fig.height = 4.5, fig.align = "center", out.width = "100%"}
plot_mix_fitness(fitness_smith_2010)
```

These plots help you see which fitness measures and mixing scales are more informative or easier to analyze. For example, within-group competition among these *Myxo* strains is negatively frequency-dependent and a linear function of log mixing ratio. 

<!--
`plot_mix_fitness()` uses `log10` scales for fitness and mixing ratio, so it will return warnings when some of these values are zero---for data with only one strain, for example. 
-->

As a tool for data inspection, `plot_mix_fitness()` shows each data point individually instead of a statistical summary like the mean value with error bars. For complex datasets with many strain combinations or treatment conditions, it may be easier to inspect subsets of the data. 

```{r compare-Yurtsev, fig.width = 6.25, fig.height = 4.5, fig.align = "center", out.width = "100%"}
fitness_Yurtsev_2013 |>
subset(ampicillin == 100 & dilution == 100) |>
plot_mix_fitness()
```

This example also shows that microbimixr functions work with the R pipe `|>`. 

For the interaction between antibiotic-resistant and sensitive *E. coli* strains shown above, total group fitness is just a constant determined by experimental conditions---how much the culture was diluted when passaged. So the within-group fitness ratio would by itself be a succinct measure of the interesting effects. 

<!--
microbimixr's plotting functions can use fitness datasets that aren't made by `calculate_mix_fitness()`, but you may need to specify what the columns mean using the `var_names` option. 
-->


## Plot fitness with microbimixr functions

<!-- 
Once you know which fitness measures to focus on, you can visualize them with microbimixr's built-in plotting functions: 

microbimixr offers several functions for plotting specific fitness measures as a function of their initial frequency 
-->

Once you know which fitness measures to focus on, you can use microbimixr to visualize them. The simplest way is to use its built-in plot functions:

* `plot_strain_fitness()`: Plot fitness of each strain separately
* `plot_total_group_fitness()` : Plot fitness of total group or subpopulation
* `plot_within_group_fitness()` : Plot within-group ratio of strain fitnesses

These functions let you reproduce the individual parts of the diagnostic plots. They're an easy way to plot fitness data with reasonable default settings. 
<!-- They work best with simple mix experiments with two strains and no other treatment variables.  -->

```{r plot-smith2010, fig.width = 2.25, fig.height = 1.75, fig.align = "center", out.width = "50%"}
plot_total_group_fitness(fitness_smith_2010)
```

They also include some basic plotting options:

```{r plot-options-smith2010, fig.width = 5.5, fig.height = 2.25, fig.align = "center", out.width = "100%"}
fig_total <- plot_total_group_fitness(
	fitness_smith_2010,
	xlab = "Initial frequency of evolved strain",
	ylab = "Total sporulation efficiency\n(spores/cell)"
)
fig_within <- plot_within_group_fitness(
	fitness_smith_2010, 
	mix_scale = "ratio",
	xlab = "Initial ratio evolved / ancestral",
	ylab = "Relative efficiency\nevolved / ancestral",
	ylim = c(0.01, 100)
)
fig_total + fig_within + plot_annotation(tag_levels = "A")
```

This example also shows how to use the patchwork package to combine subplots with `+`. 

<!--
The `plot_*_fitness()` functions return ggplot2 objects, so you can extend and modify them by adding or replacing scales, themes, and so on. 
-->


## Plot fitness with ggplot2

The fitness-plotting functions use graphics package ggplot2. You can also use microbimixr's various axes and other plot components when you make ggplot2 figures on your own. These include:

* Mixing axes `scale_x_initial_fraction()` and `scale_x_initial_ratio()`
* Fitness axes `scale_y_fitness()`, `scale_y_fitness_total()`, and `scale_y_fitness_ratio()`
* Data points `geom_point_overlap()` that easier to read when they overlap

These functions return custom plot elements for ggplot2 that have default settings appropriate for microbial mix experiments. 

```{r elements-smith, fig.width = 3, fig.height = 2.25, fig.align = "center", out.width = "50%"}
fitness_smith_2010 |>
	ggplot(aes(x = initial_fraction_A, y = fitness_total)) +
	scale_x_initial_fraction(name = "Initial frequency of evolved strain") +
	scale_y_fitness_total(name = "Sporulation efficiency\n(spores/cell)", limits = c(1e-8, 1)) +
	geom_smooth(method = "lm", se = FALSE, formula = y ~ poly(x, 2)) +
	geom_point_overlap(shape = 23, fill = "lightblue", color = "darkblue", size = 2) +
	theme_bw()
```

The plot elements are especially useful for more-complex figures with different experimental conditions or multiple strain combinations.

```{r elements-Yurtsev, fig.width = 6.25, fig.height = 2.25, out.width = "100%", align = "center"}
names_Yurtsev <- c(name_A = "resistant", name_B = "sensitive")
name_amp <- "Ampicillin\n(\u03BCg/mL)"
fitness_Yurtsev_2013 |>
	subset(
		ampicillin %in% c(0, 15, 50, 100) & dilution %in% c(200, 400, 800) &
		fitness_ratio_A_B > 0 & is.finite(fitness_ratio_A_B)
	) |>
	ggplot(mapping = aes(
		x = initial_fraction_A, y = fitness_ratio_A_B, 
		fill = factor(ampicillin), color = factor(ampicillin))
	) +
	geom_point_overlap(na.rm = TRUE) +
	scale_x_initial_fraction(strain_names = names_Yurtsev) +
	scale_y_fitness_ratio(strain_names = names_Yurtsev, limits = c(1e-1, 1e4)) +
	scale_fill_viridis_d(name = name_amp, option = "magma", direction = -1, begin = 0.5, end = 1) +
	scale_color_viridis_d(name = name_amp, option = "magma", direction = -1, begin = 0, end = 0.8) +
	facet_wrap(~ dilution, labeller = as_labeller(function(x) paste0(x, "-fold dilution"))) 
```


## References

* smith j, Van Dyken JD, and Zee PC (2010) A generalization of Hamilton's rule for the evolution of microbial cooperation. *Science* **328:** 1700-1703. 
[https://doi.org/10.1126/science.1189675](https://doi.org/10.1126/science.1189675)

* smith j and Inglis RF (2021) Evaluating kin and group selection as tools for 
quantitative analysis of microbial data. *Proceedings B* **288:** 20201657.
[https://doi.org/10.1098/rspb.2020.1657](https://doi.org/10.1098/rspb.2020.1657)

* Yurtsev EA, Chao HX, Datta MS, Artemova T, and Gore J (2013) Bacterial cheating drives the population dynamics of cooperative antibiotic resistance plasmids. *Molecular Systems Biology* **9:** 683. 
[https://doi.org/10.1038/msb.2013.39](https://doi.org/10.1038/msb.2013.39)
