---
title: "Analyzing microbial interactions with microbimixr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing microbial interactions with microbimixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(microbimixr)
library(patchwork)
```

A common experimental design in microbiology is to mix together two different 
microbes (different genotypes of bacteria, for example) then measure how their 
behavior depends on mix frequency. 
How do they act differently together than when they're on their own? 
How does mixing affect their survival and reproductive success (their fitness)?

microbimixr is an R package for analyzing these mix experiments. 
It helps researchers get the most out of their data by providing convenience functions to 
calculate and plot fitness measures that are: 

* Robust and quantitatively comparable across different species and types of 
  interaction
* Meaningful for both kin and multilevel selection theories of social evolution
* Well-suited to statistical analysis of effect sizes and confidence intervals


<!-- TEACH PEOPLE HOW TO USE THE PACKAGE -->

The basic workflow for microbimixr is: 

1. Make an input data frame containing the experimental observations used to calculate 
   fitness
2. Calculate fitness measures with `calculate_mix_fitness()`
3. Inspect a diagnostic plot of fitness measures with `plot_mix_fitness()`
4. Plot measures of interest with `plot_strain_fitness()`, 
  `plot_total_group_fitness()`, and/or `plot_within_group_fitness()`
5. Fit statistical models of your choosing and plot them with the data


## Input data

The input format for microbimixr is a data frame where each row describes microbial abundance at the beginning and end of a mix experiment. The experimentally observed quantities could be any combination of strain abundance (number or density), strain frequency (fraction of total), or total group abundance---anything sufficient to define the absolute abundance of both strains in the experiment. Each row should contain data for a single observational unit: a single experimental replicate with one strain combination, one mix frequency, one combination of experimental treatments, etc.  

For example, let's look at some datasets included with microbimixr. This dataset from smith *et al.* (2010) is an experiment mixing a wild-type strain of *Myxococcus* bacteria with an experimentally-evolved cheater strain. The observed quantities are the number of cells from each strain before and after they form multicellular fruiting bodies. 

```{r data-smith-2010}
head(data_smith_2010)
```

This dataset from Yurtsev *et al.* (2013) describes the results of mix experiments with an antibiotic-sensitive strain of *Escherichia coli* and an antibiotic-resistant strain that detoxifies its local environment. The observed quantities here are total cell density (from OD~600~) and the fraction of cells belonging to each strain (from flow cytometry). The dataset also includes several experimental treatments with different antibiotic concentrations and different amounts of culture growth. 

```{r data-Yurtsev-2013}
head(data_Yurtsev_2013)
```
* Explain how made to take tidy data, best archiveable form


## Calculating fitness

The microbimixr function to calculate fitness is `calculate_mix_fitness()`. It accepts an input data frame of microbial abundances and a character vector describing what the data columns are. It returns a data frame with several calculated measures of fitness and two measures of mix frequency. For example: 

```{r fitness_smith_2010}
fitness_smith_2010 <- calculate_mix_fitness(
	data = data_smith_2010, 
	var_names = c(
		initial_number_A = "initial_cells_evolved",
		initial_number_B = "initial_cells_ancestral",
		final_number_A = "final_spores_evolved",
		final_number_B = "final_spores_ancestral", 
		name_A = "GVB206.3", 
		name_B = "GJV10"
	)
)
head(fitness_smith_2010)
```

To include data columns in the fitness output that describe experimental treatments, experimental block, and so on, use the `keep` argument. For the Yurtsev data, for example, we want to keep antibiotic concentration and culture dilution: 

```{r fitness_Yurtsev_2013}
fitness_Yurtsev_2013 <- calculate_mix_fitness(
	data_Yurtsev_2013, 
	var_names = c(
		initial_number_total = "OD_initial",
		initial_fraction_A = "fraction_resistant_initial",
		final_number_total = "OD_final",
		final_fraction_A = "fraction_resistant_final",
		name_A = "AmpR",
		name_B = "AmpS"
	), 
	keep = c("ampicillin", "dilution")
)
head(fitness_Yurtsev_2013)
```

Note the warning messages here. `calculate_mix_fitness()` will warn you if any of the data are not biologically meaningful. There cannot be a negative number of cells, for example, and fractions are bounded at zero and one. Some of the Yurtsev data include negative fractions---likely a minor artifact of correcting for background fluorescence in flow cytometry. 

<!-- 
* Explain `name_A` and `name_B` arguments?
* They make explicit which is which in the output itself
* Can be column or label
* Column example: Madgwick data
-->


## Fitness math

`calculate_mix_fitness()` calculates four fitness outcomes from the input data:

* `fitness_A` : Fitness of strain A
* `fitness_B` : Fitness of strain B
* `fitness_total` : Total-group fitness (strains A and B combined)
* `fitness_ratio_A_B` : Within-group relative fitness measured as `fitness_A / fitness_B`

Some interaction effects are easier to visualize, analyze, and interpret with strain-focused fitness outcomes (`fitness_A` and `fitness_B`), while others are better suited to group-focused outcomes (`fitness_total` and `fitness_ratio_A_B`). So microbimixr includes both approaches by default. 

The actual calculations are very simple. If the initial abundance of strain A (measured as the number or density of cells or virions) is $n_A$ and its final abundance is $n'_A$, then its fitness is $w_A = n'_A/n_A$. Similarly, the fitness of strain B is $w_B = n'_B/n_B$. Total group fitness is $W = (n'_A + n'_B)/(n_A + n_B)$ and the within-group fitness ratio is $w_A/w_B$. 

These values are unscaled Wrightian fitness measured over the entire time period between the initial and final abundances. They measure fold change in absolute abundance. For example: if a bacterial strain's cell density increases 100-fold, its fitness will be $w = 100$. If it decreases to 10% of its initial density, its fitness will be $w = 0.1$.

These fitness measures are robust across microbial species and types of interaction, make fitness effects quantitatively comparable across systems, and can be meaningfully incorporated into theoretical models of microbial social evolution (smith & Inglis 2021). They are usually best visualized and analyzed over $\log_{10}$ scales. 

`calculate_mix_fitness()` also calculates two measures of mixing frequency: 

* `initial_fraction_A` : Initial frequency of strain A measured as the fraction of all cells or virions
* `initial_ratio_A_B` : Initial ratio of strain A to strain B

Microbial interactions may be more relevant or more linear over one scale than the other, so microbimixr includes both by default. `initial_fraction_A` is calculated as $n_A / (n_A + n_B)$ and runs from 0 to 1. `initial_ratio_A_B` is calculated as $n_A / n_B$. It runs from negative infinity to positive infinity and is best analyzed over $\log_{10}$ scales. 


## Diagnostic plots

To quickly see an overview of the various fitness effects in a dataset, use `plot_mix_fitness()`. 
<!-- `plot_mix_fitness()` produces a diagnostic overview of the various fitness effects in a dataset.  -->
It draws a combined plot of strain, total group, and relative within-group fitness against two measures of mix frequency.

```{r fig.width = 6, fig.height = 4.5, fig.align = "center", out.width = "100%"}
plot_mix_fitness(fitness_smith_2010)
```

These plots help you quickly see which fitness measures and mix frequency scales are more informative or easier to analyze. For example, within-group competition among these *Myxo* strains is negatively frequency-dependent and a linear function of log mixing ratio. 

`plot_mix_fitness()` uses `log10` scales for fitness and mixing ratio, so it will return warnings when some of these values are zero---for data with only one strain, for example. 

As a tool for data inspection, `plot_mix_fitness()` shows each data point individually instead of a statistical summary like the mean value with error bars. For complex datasets with many strain combinations or treatment conditions, it may be easier to inspect subsets of the data. microbimixr functions work with the R pipe `|>`. 

```{r fig.width = 6, fig.height = 4.5, fig.align = "center", out.width = "100%"}
fitness_Yurtsev_2013 |>
subset(ampicillin == 100 & dilution == 100) |>
plot_mix_fitness()
```

For this interaction between antibiotic-resistant and sensitive *E. coli* strains, total group fitness is just a constant determined by experimental conditions (how much the culture was diluted when passaged). So the within-group fitness ratio would by itself be a succinct measure of the interesting effects. 

<!--
microbimixr's plotting functions can use fitness datasets that aren't made by `calculate_mix_fitness()`, but you may need to specify what the columns mean using the `var_names` option. 
-->


## Plotting specific fitness measures

Once you know which fitness measures to focus on, you can visualize them with microbimixr's built-in plotting functions: 

<!-- microbimixr offers several functions for plotting specific fitness measures as a function of their initial frequency -->

* `plot_strain_fitness()`: Plots absolute fitness for each of two strains
* `plot_total_group_fitness()` : Plots total-group fitness
* `plot_within_group_fitness()` : Plots relative within-group fitness of two strains

These are convenience functions that use the ggplot2 graphics package to plot fitness data with reasonable default settings. 

```{r fig.width = 2.75, fig.height = 2.25, fig.align = "center", out.width = "50%"}
plot_total_group_fitness(fitness_smith_2010)
```

The default mixing scale for the `plot_*_fitness()` is the initial proportion of strain A. To instead use the ratio of strain A to strain B (on a log~10~ scale), use the `mix_scale = "ratio"` option: 

```{r fig.width = 2.75, fig.height = 2.25, fig.align = "center", out.width = "50%"}
plot_within_group_fitness(fitness_smith_2010, mix_scale = "ratio")
```

* Demonstrate functions and options
* Describe default setting choices: scales, etc

The `plot_*_fitness()` functions return ggplot2 objects, so you can extend and modify them by adding or replacing scales, themes, and so on. 

```
code_example_here
```

* Panels/colors for mix expts under different conditions (Gore data)
* Combine subplots with patchwork package?

As your results and figures become more complex, you may find it simpler to just make them with ggplot2 directly, from the ground up. 

* Make way to reuse microbimixr scales in direct `ggplot()` construction?


## Plotting data with fitted statistical models

...


## References

* smith j, Van Dyken JD, and Zee PC (2010) A generalization of Hamilton's rule for the evolution of microbial cooperation. *Science* **328:** 1700-1703. 
[https://doi.org/10.1126/science.1189675](https://doi.org/10.1126/science.1189675)

* smith j and Inglis RF (2021) Evaluating kin and group selection as tools for 
quantitative analysis of microbial data. *Proceedings B* **288:** 20201657.
[https://doi.org/10.1098/rspb.2020.1657](https://doi.org/10.1098/rspb.2020.1657)

* Yurtsev EA, Chao HX, Datta MS, Artemova T, and Gore J (2013) Bacterial cheating drives the population dynamics of cooperative antibiotic resistance plasmids. *Molecular Systems Biology* **9:** 683. 
[https://doi.org/10.1038/msb.2013.39](https://doi.org/10.1038/msb.2013.39)
